---
title: "betAS tutorial"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
    %\VignetteIndexEntry{Case study: command-line interface (CLI) tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r}

library(betAS)
library(ggplot2)

```

Load files and variables

```{r}

sampleTable           <- readRDS("samplesTable.rds")
maxDevSimulationN100  <- readRDS("xintercepts_100incr_100cov_100trials.rds")
pastelColors          <- c("#FF9AA2", "#FFB7B2", "#FFDAC1", "#E2F0CB", "#B5EAD7", "#C7CEEA", "#FBE2FD", "#D9ECFE")

```

# Simulate loading a (vast-tools) table into the app
Demo file: only exon skipping event types ("C1", "C2", "C3", "S" and "MIC"), minimum coverage without NAs

```{r}

testTable   <- readRDS("testTableEX.rds")
cat(paste0("Filtered events: ", nrow(testTable$PSI)))
table(testTable$PSI$COMPLEX)

```

# (optional) Select "alternative" events (1 < PSI < 99)

```{r}

testTable <- alternativeVastTools(testTable, minPsi = 1, maxPsi = 99)
cat(paste0("Alternative events: ", nrow(testTable$PSI)))

```

# Define PSI and Qual table to be used in the following sections

```{r}

psiObject <- testTable
psiTable  <- psiObject$PSI
qualTable <- psiObject$Qual

```

# "Big picture" plot

```{r}

bigPicturePlot <- bigPicturePlot(table = psiTable)
bigPicturePlot + theme_minimal()

```

# Define groups based on sampleTable

```{r}

groups <- unique(sampleTable[,"organism_part"])
random_colors <- pastelColors
groupList <- list()

for(i in 1:length(groups$organism_part)){

  groupNames <- sampleTable$Run[which(sampleTable[,"organism_part"] == groups$organism_part[i])]

  # Assign new group
  currentNames <- names(groupList)
  groupList[[length(groupList)+1]] <- list(name = groups$organism_part[i],
                                           samples = groupNames,
                                           color = random_colors[1])
  names(groupList) <- make.unique(c(currentNames, groups$organism_part[i]))

  random_colors <- random_colors[-1]

}

```

# Plot densities for defined groups/samples

```{r}

tdensities <- plotIndividualDensitiesList(eventID = "HsaEX0019479",
                                          npoints = 500,
                                          psitable = psiTable,
                                          qualtable = qualTable,
                                          groupList = groupList,
                                          maxDevTable = maxDevSimulationN100)

tdensities + theme_minimal()

```

# Perform betAS

```{r}

groupA    <- "heart"
groupB    <- "forebrain"

samplesA    <- groupList[[groupA]]$samples
samplesB    <- groupList[[groupB]]$samples

colsGroupA    <- convertCols(psiTable, samplesA)
colsGroupB    <- convertCols(psiTable, samplesB)

```

## Volcano plot with Pdiff as y-axis variable

```{r}

volcanoTable <- prepareTableVolcano(psitable = psiTable,
                                    qualtable = qualTable,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100)

volcanoPdiff <- plotVolcano(betasTable = volcanoTable,
                            labA = groupList$heart,
                            labB = groupList$forebrain,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C")

volcanoPdiff + theme_minimal()

```

