---
title: "betAS command-line interface (CLI) tutorial"
author: "Mariana Ascensão-Ferreira"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
    %\VignetteIndexEntry{betAS command-line interface (CLI) tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

*betAS* is a user-friendly R package that allows intuitive analysis and
visualisation of differential alternative splicing (AS) based on
beta distributions.

Beta distributions are suitable to quantify inclusion proportions of
alternative sequences, using RNA sequencing reads supporting their
inclusion and exclusion as surrogates for the two distribution shape
parameters. Each such beta distribution has the inclusion proportion as
mean value and is narrower when the read coverage is higher, facilitating
the interpretability of its precision when plotted. betAS uses beta
distributions to accurately model PSI values and their precision, to
quantitatively and visually compare AS between groups of samples.

betAS allows the analysis of user-provided tables with AS quantifications,
such as those obtained by vast-tools (https://github.com/vastgroup/vast-tools),
ranking differentially spliced events by a significance metric that
incorporates the compromise between the uncertainty in individual sample
estimates and the variability among replicates.

# Installing and starting the program
Install *betAS* by typing the following in an R console:

```{r install, eval=FALSE}

install.packages("BiocManager")
BiocManager::install("betAS")

```

After the installation, load betAS and other required packages by typing:

```{r load, message=FALSE}

library(betAS)
library(ggplot2)

```

# Case study: differential alternative splicing analysis between two groups
Load required files and variables

```{r}

sampleTable           <- readRDS("samplesTable.rds")
maxDevSimulationN100  <- readRDS("xintercepts_100incr_100cov_100trials.rds")
pastelColors          <- c("#FF9AA2", "#FFB7B2", "#FFDAC1", "#E2F0CB", "#B5EAD7", "#C7CEEA", "#FBE2FD", "#D9ECFE")

```

## Simulate loading a (vast-tools) table into the app
Demo file: only exon skipping event types ("C1", "C2", "C3", "S" and "MIC"), minimum coverage without NAs

```{r}

testTable   <- readRDS("testTableEX.rds")
cat(paste0("Filtered events: ", nrow(testTable$PSI)))
table(testTable$PSI$COMPLEX)

```

## (optional) Select "alternative" events (1 < PSI < 99)

```{r}

testTable <- alternativeVastTools(testTable, minPsi = 1, maxPsi = 99)
cat(paste0("Alternative events: ", nrow(testTable$PSI)))

```

## Define PSI and Qual table to be used in the following sections

```{r}

psiObject <- testTable
psiTable  <- psiObject$PSI
qualTable <- psiObject$Qual

```

## "Big picture" plot

```{r}

bigPicturePlot <- bigPicturePlot(table = psiTable)
bigPicturePlot + theme_minimal()

```

## Define groups based on sampleTable

```{r}

groups <- unique(sampleTable[,"organism_part"])
random_colors <- pastelColors
groupList <- list()

for(i in 1:length(groups$organism_part)){

  groupNames <- sampleTable$Run[which(sampleTable[,"organism_part"] == groups$organism_part[i])]

  # Assign new group
  currentNames <- names(groupList)
  groupList[[length(groupList)+1]] <- list(name = groups$organism_part[i],
                                           samples = groupNames,
                                           color = random_colors[1])
  names(groupList) <- make.unique(c(currentNames, groups$organism_part[i]))

  random_colors <- random_colors[-1]

}

```

## Plot densities for defined groups/samples

```{r}

tdensities <- plotIndividualDensitiesList(eventID = "HsaEX0019479",
                                          npoints = 500,
                                          psitable = psiTable,
                                          qualtable = qualTable,
                                          groupList = groupList,
                                          maxDevTable = maxDevSimulationN100)

tdensities + theme_minimal()

```

# Apply betAS

```{r}

groupA    <- "heart"
groupB    <- "forebrain"

samplesA    <- groupList[[groupA]]$samples
samplesB    <- groupList[[groupB]]$samples

colsGroupA    <- convertCols(psiTable, samplesA)
colsGroupB    <- convertCols(psiTable, samplesB)

```

## Volcano plot with Pdiff as y-axis variable

```{r}

volcanoTable <- prepareTableVolcano(psitable = psiTable,
                                    qualtable = qualTable,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100)

volcanoPdiff <- plotVolcano(betasTable = volcanoTable,
                            labA = groupList$heart,
                            labB = groupList$forebrain,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C")

volcanoPdiff + theme_minimal()

```

# Session information

```{r si}

sessionInfo()

```

